---
title: "CDC ILI forecast"
output: 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We used New York State ILI counts as an example to demonstrate phylodynamics forecasting. To obtain latest data from CDC, we install the `cdcfluview` package:

```r
devtools::install_github("hrbrmstr/cdcfluview")
```

We first visualized the total ILI counts for each flu season (week 40 - week 21). 

```{r}
library(cdcfluview)
library(dplyr)
library(ggplot2)
library(phylodyn)
library(PILAF)

selected_seasons <- c("2014-2015", "2015-2016", "2016-2017", "2017-2018", "2018-2019")

NY_ILI <- ilinet("state") %>%
  filter(region == "New York") %>%
  select("year", "week", "ilitotal", "week_start") %>%
  rename(ILI_total = ilitotal)

dates <- readr::read_tsv(system.file("extdata/H3N2/NewYork_A_H3N2_20190307_filtered.tsv",
                            package = "PILAF"))
min(dates$collection_date)
max(dates$collection_date)

dates <- readr::read_tsv(system.file("extdata/H1N1/NewYork_A_H1N1_20190318_filtered.tsv",
                                     package = "PILAF"))
min(dates$collection_date)
max(dates$collection_date)

# use 2019-02-04 as end date
NY_ILI <- NY_ILI %>%
  filter(week_start <= as.Date("2019-02-04"))

p_ILI <- visualize_flu_season(NY_ILI, time_series_names = "ILI_total", 
                              subset_seasons = selected_seasons, ncol = 5, season_label = F)
p_ILI
```

We can also visualize the inferred trajectory. 

```{r H3N2_tree}
H3N2_tree <- ape::read.nexus(system.file("extdata/H3N2/MCC.tree",
                                         package = "PILAF"))
H3N2_last_time <- 2019.0739726027398
# AR(2) + cRW(1) model
ar2crw1_formula <- y ~ -1 + f(time, model = "ar", order = 2) +
      f(week, model = "rw1", cyclic = T, constr = F)
H3N2_Ne_forecast <- BNPR_forecast(H3N2_tree, last_time = H3N2_last_time, formula = ar2crw1_formula, pred = 0)
H3N2_Ne_data <- with(H3N2_Ne_forecast, 
                     data.frame(time = H3N2_last_time - result$summary.random$time$ID,
                                week = weeks,
                                H3N2_Ne = effpopmean,
                                lwr = effpop025,
                                upr = effpop975))
H3N2_Ne_data$year <- floor(H3N2_Ne_data$time)
p_H3N2_Ne <- visualize_flu_season(H3N2_Ne_data, time_series_names = "H3N2_Ne", 
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H3N2_Ne

# BNPR model
H3N2_BNPR <- BNPR(H3N2_tree, lengthout = 300)
H3N2_BNPR_data <- with(H3N2_BNPR, 
                     data.frame(time = H3N2_last_time - result$summary.random$time$ID,
                                H3N2_Ne = effpopmean,
                                lwr = effpop025,
                                upr = effpop975))
H3N2_BNPR_data$week <- lubridate::week(lubridate::date_decimal(H3N2_BNPR_data$time))
H3N2_BNPR_data$year <- floor(H3N2_BNPR_data$time)
p_H3N2_BNPR <- visualize_flu_season(H3N2_BNPR_data, time_series_names = "H3N2_Ne", 
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H3N2_BNPR


# BNPR PS model
H3N2_BNPR_PS <- BNPR_PS(H3N2_tree, lengthout = 300)
H3N2_BNPR_PS_data <- with(H3N2_BNPR_PS, 
                     data.frame(time = H3N2_last_time - result$summary.random$time$ID,
                                H3N2_Ne = effpopmean,
                                lwr = effpop025,
                                upr = effpop975))
H3N2_BNPR_PS_data$week <- lubridate::week(lubridate::date_decimal(H3N2_BNPR_PS_data$time))
H3N2_BNPR_PS_data$year <- floor(H3N2_BNPR_PS_data$time)
p_H3N2_BNPR_PS <- visualize_flu_season(H3N2_BNPR_PS_data, time_series_names = "H3N2_Ne", 
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H3N2_BNPR_PS

# seaonal model
seasonal_formula <- y ~ f(time, model="ar", order = 2) + f(seasonal, model="seasonal",
season.length = 52)
H3N2_seasonal_forecast <- BNPR_forecast(H3N2_tree, last_time = H3N2_last_time, formula = seasonal_formula, pred = 0)
H3N2_seasonal_data <- with(H3N2_seasonal_forecast, 
                     data.frame(time = H3N2_last_time - result$summary.random$time$ID,
                                week = weeks,
                                H3N2_Ne = effpopmean,
                                lwr = effpop025,
                                upr = effpop975))
H3N2_seasonal_data$year <- floor(H3N2_seasonal_data$time)
p_H3N2_seasonal <- visualize_flu_season(H3N2_seasonal_data, time_series_names = "H3N2_Ne", 
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H3N2_seasonal
```

```{r H1N1_tree}
H1N1_tree <- ape::read.nexus(system.file("extdata/H1N1/MCC.tree",
                                         package = "PILAF"))
H1N1_last_time <- 2019.0931506849315

# AR(2) + cRW(1)
ar2crw1_formula <- y ~ -1 + f(time, model = "ar", order = 2) +
      f(week, model = "rw1", cyclic = T, constr = F)
H1N1_Ne_forecast <- BNPR_forecast(H1N1_tree, last_time = H1N1_last_time, formula = ar2crw1_formula, pred = 0)
H1N1_Ne_data <- with(H1N1_Ne_forecast, data.frame(time = H1N1_last_time - result$summary.random$time$ID,
                                             week = weeks,
                                             H1N1_Ne = effpopmean,
                                             lwr = effpop025,
                                             upr = effpop975
                                             ))
H1N1_Ne_data$year <- floor(H1N1_Ne_data$time)
p_H1N1_Ne <- visualize_flu_season(H1N1_Ne_data, time_series_names = "H1N1_Ne",
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H1N1_Ne

# BNPR
H1N1_BNPR <- BNPR(H1N1_tree, lengthout = 300)
H1N1_BNPR_data <- with(H1N1_BNPR, 
                     data.frame(time = H1N1_last_time - result$summary.random$time$ID,
                                H1N1_Ne = effpopmean,
                                lwr = effpop025,
                                upr = effpop975))
H1N1_BNPR_data$week <- lubridate::week(lubridate::date_decimal(H1N1_BNPR_data$time))
H1N1_BNPR_data$year <- floor(H1N1_BNPR_data$time)
p_H1N1_BNPR <- visualize_flu_season(H1N1_BNPR_data, time_series_names = "H1N1_Ne", 
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H1N1_BNPR

# BNPR-PS
H1N1_BNPR_PS <- BNPR_PS(H1N1_tree, lengthout = 300)
H1N1_BNPR_PS_data <- with(H1N1_BNPR_PS, 
                     data.frame(time = H1N1_last_time - result$summary.random$time$ID,
                                H1N1_Ne = effpopmean,
                                lwr = effpop025,
                                upr = effpop975))
H1N1_BNPR_PS_data$week <- lubridate::week(lubridate::date_decimal(H1N1_BNPR_PS_data$time))
H1N1_BNPR_PS_data$year <- floor(H1N1_BNPR_PS_data$time)
p_H1N1_BNPR_PS <- visualize_flu_season(H1N1_BNPR_PS_data, time_series_names = "H1N1_Ne", 
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H1N1_BNPR_PS

# seasonal model
seasonal_formula <- y ~ f(time, model="ar", order = 2) + f(seasonal, model="seasonal",
season.length = 52)
H1N1_seasonal_forecast <- BNPR_forecast(H1N1_tree, last_time = H1N1_last_time, formula = seasonal_formula, pred = 0)
H1N1_seasonal_data <- with(H1N1_seasonal_forecast, 
                     data.frame(time = H1N1_last_time - result$summary.random$time$ID,
                                week = weeks,
                                H1N1_Ne = effpopmean,
                                lwr = effpop025,
                                upr = effpop975))
H1N1_seasonal_data$year <- floor(H1N1_seasonal_data$time)
p_H1N1_seasonal <- visualize_flu_season(H1N1_seasonal_data, time_series_names = "H1N1_Ne", 
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H1N1_seasonal
```

```{r}
Yamagata_tree <- ape::read.nexus(system.file("extdata/Yamagata/MCC.tree",
                                         package = "PILAF"))
Yamagata_last_time <- 2019.0739726027398
ar2crw1_formula <- y ~ -1 + f(time, model = "ar", order = 2) +
      f(week, model = "rw1", cyclic = T, constr = F)
Yamagata_Ne_forecast <- BNPR_forecast(Yamagata_tree, last_time = Yamagata_last_time, formula = ar2crw1_formula, pred = 0)
Yamagata_Ne_data <- with(Yamagata_Ne_forecast, data.frame(time = Yamagata_last_time - result$summary.random$time$ID,
                                             week = weeks,
                                             Yamagata_Ne = effpopmean,
                                             lwr = effpop025,
                                             upr = effpop975))

Yamagata_Ne_data$year <- floor(Yamagata_Ne_data$time)
p_Yamagata_Ne <- visualize_flu_season(Yamagata_Ne_data, time_series_names = "Yamagata_Ne",
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_Yamagata_Ne
```

```{r}
Victoria_tree <- ape::read.nexus(system.file("extdata/Victoria/MCC.tree",
                                         package = "PILAF"))
Victoria_last_time <- 2018.317808219178
ar2crw1_formula <- y ~ -1 + f(time, model = "ar", order = 2) +
      f(week, model = "rw1", cyclic = T, constr = F)
Victoria_Ne_forecast <- BNPR_forecast(Victoria_tree, last_time = Victoria_last_time, formula = ar2crw1_formula, pred = 0)
Victoria_Ne_data <- with(Victoria_Ne_forecast, data.frame(time = Victoria_last_time - result$summary.random$time$ID,
                                             week = weeks,
                                             Victoria_Ne = effpopmean,
                                             lwr = effpop025,
                                             upr = effpop975))
Victoria_Ne_data$year <- floor(Victoria_Ne_data$time)
p_Victoria_Ne <- visualize_flu_season(Victoria_Ne_data, time_series_names = "Victoria_Ne",
                                      subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                      error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_Victoria_Ne
```

We can alsoview each season separately. 

```{r include = F}
theme_no_x <-  theme(
  axis.title.x = element_blank(), 
  axis.line.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.text.x = element_blank())
p_ILI <- p_ILI + ylab("ILI total") + theme_no_x
p_H3N2_BNPR <- p_H3N2_BNPR + ylab("H3N2 Ne") + theme_no_x + theme(strip.text = element_blank())
p_H1N1_BNPR <- p_H1N1_BNPR + ylab("H1N1 Ne") + theme_no_x + theme(strip.text = element_blank())
p_H3N2_BNPR_PS <- p_H3N2_BNPR_PS + ylab("H3N2 Ne (PS)") + theme_no_x + theme(strip.text = element_blank())
p_H1N1_BNPR_PS <- p_H1N1_BNPR_PS + ylab("H1N1 Ne (PS)") + theme(strip.text = element_blank())

p2 <- cowplot::plot_grid(p_ILI, p_H3N2_BNPR, p_H3N2_BNPR_PS, p_H1N1_BNPR, p_H1N1_BNPR_PS, ncol = 1, align = "v")
ggsave(p2, file = "../../projectingvirus/manuscript/Figures/data/Ne_ILI.pdf", width = 9, height = 8)

p_H3N2_BNPR <- p_H3N2_BNPR + ylab("H3N2 Ne") + theme_no_x + theme(strip.text = element_blank())
p_H1N1_BNPR <- p_H1N1_BNPR + ylab("H1N1 Ne") + theme(strip.text = element_blank())
p3 <- cowplot::plot_grid(p_ILI, p_H3N2_BNPR, p_H1N1_BNPR, ncol = 1, align = "v")
ggsave(p3, file = "../../projectingvirus/manuscript/Figures/data/Ne_ILI_BNPR.pdf", width = 9, height = 5)

p_H3N2_seasonal <- p_H3N2_seasonal + ylab("H3N2 Ne") + theme_no_x + theme(strip.text = element_blank())
p_H1N1_seasonal <- p_H1N1_seasonal + ylab("H1N1 Ne") + theme(strip.text = element_blank())
p4 <- cowplot::plot_grid(p_ILI, p_H3N2_seasonal, p_H1N1_seasonal, ncol = 1, align = "v")
ggsave(p4, file = "../../projectingvirus/manuscript/Figures/data/Ne_ILI_seasonal.pdf", width = 9, height = 5)

p <- cowplot::plot_grid(p3, p4, ncol = 1, labels = c("a", "b"))
ggsave(p, file = "../../projectingvirus/manuscript/Figures/data/Ne_ILI_all.pdf", width = 9, height = 10)
```

cRW(1) + AR(2): Quantify the similarity between time series using Spearman correlation.

```{r}
joined_ts1 <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H3N2_Ne_data, year, week, H3N2_Ne), by = c("year", "week"))  %>%
  select(ILI_total, H3N2_Ne) 
cor.test(joined_ts1$ILI_total, joined_ts1$H3N2_Ne, method = "pearson")
joined_ts2 <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H1N1_Ne_data, year, week, H1N1_Ne), by = c("year", "week")) %>%
  select(ILI_total, H1N1_Ne) 
cor.test(joined_ts2$ILI_total, joined_ts2$H1N1_Ne, method = "pearson")
joined_ts <- select(H3N2_Ne_data, year, week, H3N2_Ne) %>%
  inner_join(select(H1N1_Ne_data, year, week, H1N1_Ne), by = c("year", "week"))
cor.test(joined_ts$H3N2_Ne, joined_ts$H1N1_Ne, method = "pearson")
joined_all <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H3N2_Ne_data, year, week, H3N2_Ne), by = c("year", "week")) %>%
  dplyr::inner_join(select(H1N1_Ne_data, year, week, H1N1_Ne), by = c("year", "week")) %>%
  select(ILI_total, H3N2_Ne, H1N1_Ne)
lm_res <- lm(ILI_total ~ H3N2_Ne + H1N1_Ne, data = joined_all)
summary(lm_res)
```

BNPR: Regression analysis between ILI and the two virus time series.

```{r}
joined_ts1 <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H3N2_BNPR_data, year, week, H3N2_Ne), by = c("year", "week"))  %>%
  select(ILI_total, H3N2_Ne) 
cor.test(joined_ts1$ILI_total, joined_ts1$H3N2_Ne, method = "pearson")
joined_ts2 <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H1N1_BNPR_data, year, week, H1N1_Ne), by = c("year", "week")) %>%
  select(ILI_total, H1N1_Ne) 
cor.test(joined_ts2$ILI_total, joined_ts2$H1N1_Ne, method = "pearson")
joined_ts <- select(H3N2_BNPR_data, year, week, H3N2_Ne) %>%
  inner_join(select(H1N1_BNPR_data, year, week, H1N1_Ne), by = c("year", "week"))
cor.test(joined_ts$H3N2_Ne, joined_ts$H1N1_Ne, method = "pearson")
joined_all <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H3N2_BNPR_data, year, week, H3N2_Ne), by = c("year", "week")) %>%
  dplyr::inner_join(select(H1N1_BNPR_data, year, week, H1N1_Ne), by = c("year", "week")) %>%
  select(ILI_total, H3N2_Ne, H1N1_Ne)
lm_res <- lm(ILI_total ~ H3N2_Ne + H1N1_Ne, data = joined_all)
summary(lm_res)
```

Comparison with BNPR PS

```{r}
joined_ts1 <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H3N2_BNPR_PS_data, year, week, H3N2_Ne), by = c("year", "week"))  %>%
  select(ILI_total, H3N2_Ne) 
cor.test(joined_ts1$ILI_total, joined_ts1$H3N2_Ne, method = "pearson")
joined_ts2 <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H1N1_BNPR_PS_data, year, week, H1N1_Ne), by = c("year", "week")) %>%
  select(ILI_total, H1N1_Ne) 
cor.test(joined_ts2$ILI_total, joined_ts2$H1N1_Ne, method = "pearson")
joined_ts <- select(H3N2_BNPR_PS_data, year, week, H3N2_Ne) %>%
  inner_join(select(H1N1_BNPR_PS_data, year, week, H1N1_Ne), by = c("year", "week"))
cor.test(joined_ts$H3N2_Ne, joined_ts$H1N1_Ne, method = "pearson")
joined_all <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H3N2_BNPR_PS_data, year, week, H3N2_Ne), by = c("year", "week")) %>%
  dplyr::inner_join(select(H1N1_BNPR_PS_data, year, week, H1N1_Ne), by = c("year", "week")) %>%
  select(ILI_total, H3N2_Ne, H1N1_Ne)
lm_res <- lm(ILI_total ~ H3N2_Ne + H1N1_Ne, data = joined_all)
summary(lm_res)
```

# composition of ILI 

```{r}
clinical_data <- cdcfluview::who_nrevss("state")
NY_public_lab <- clinical_data$public_health_labs %>%
  filter(region == "New York") # a summary for each season
NY_clinical <- clinical_data$clinical_labs %>%
  filter(region == "New York") %>% # available for each week
  mutate(percent_a = as.numeric(percent_a),
         percent_b = as.numeric(percent_b))
sd(NY_clinical$percent_a)
range(NY_clinical$percent_a)
sd(NY_clinical$percent_b)
range(NY_clinical$percent_b)
NY_clinical_percent <- NY_clinical %>%
  select(wk_date, percent_a, percent_b) %>%
  mutate(wk_date = lubridate::decimal_date(wk_date)) %>%
  tidyr::gather(type, percent, -c(wk_date)) %>%
  mutate(type = toupper(gsub("percent_", "", type))) %>%
  mutate(percent = as.numeric(percent)) 
p_NY <- ggplot(data = NY_clinical_percent, aes(x = wk_date, y = percent)) +
  geom_area(aes(fill = type), position = "stack") +
  ylab("% positive") +
  theme_classic() +
  theme(legend.title = element_blank(), axis.title.x = element_blank()) +
  scale_fill_viridis_d()
```

```{r}
head(NY_public_lab)
public_lab <- NY_public_lab %>%
  select(-c(region_type, region, wk_date)) %>%
  tidyr::gather(type, count, -c(season_description, total_specimens)) %>%
  mutate(count = as.numeric(count), total_specimens = as.numeric(total_specimens)) %>%
  group_by(season_description) %>%
  mutate(total_positive_specimen = sum(count)) %>%
  mutate(percent = count / total_positive_specimen*100) %>%
  filter(type != "h3n2v")

public_lab$season <- purrr::map_chr(public_lab$season_description, 
                                    ~strsplit(.x, split = " ")[[1]][[2]])
public_lab$season = factor(public_lab$season, levels = c("2015-16", "2016-17", "2017-18", "2018-19")) 

rename_virus <- Vectorize(function(x) {
  switch(x, 
         a_2009_h1n1 = "A/H1N1",
         a_h3 = "A/H3",
         a_subtyping_not_performed = "A",
         b = "B",
         bvic = "B/Victoria",
         byam = "B/Yamagata"
         )
})
public_lab$type = rename_virus(public_lab$type)


p <- ggplot(data = public_lab) +
  geom_area(aes(x = season, y = percent, group = type, fill = type), position = 'stack') +
  theme_classic() +
  ylab("% positive") +
  theme(legend.title = element_blank())
#ggsave(p, file = "../../projectingvirus/manuscript/Figures/data/NY_subtype.pdf",
#       width = 6, height = 3)
```


Summary statistics for New York Public Health Lab information

```{r}
head(NY_public_lab)
```



National level subtype data

```{r}
national_clinical_data <- cdcfluview::who_nrevss("national")
national_public_health <- national_clinical_data$public_health_labs %>%
  select(-c(region_type, region)) %>%
  tidyr::gather(type, count, -c(year, week, total_specimens, wk_date)) %>%
  mutate(count = as.numeric(count), total_specimens = as.numeric(total_specimens)) %>%
  mutate(percent = count / total_specimens*100) %>%
  filter(type != "h3n2v")
national_public_health$type = rename_virus(national_public_health$type)

p_national <- ggplot(data = national_public_health) +
  geom_area(aes(x = wk_date, y = percent, group = type, fill = type), position = 'stack') +
  theme_classic() +
  ylab("% positive") +
  theme(legend.title = element_blank(), axis.title.x = element_blank())
```

```{r}
p_all <- cowplot::plot_grid(p_NY, p, labels = c("a", "b"), ncol = 1)
ggsave(p_all, file = "../../projectingvirus/manuscript/Figures/data/NY_national_subtype.pdf",
       width = 6, height = 6)
```

