---
title: "CDC ILI forecast"
output: 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We used New York State ILI counts as an example to demonstrate phylodynamics forecasting. To obtain latest data from CDC, we install the `cdcfluview` package:

```r
devtools::install_github("hrbrmstr/cdcfluview")
```

We first visualized the total ILI counts for each flu season (week 40 - week 21). 

```{r}
library(cdcfluview)
library(dplyr)
library(ggplot2)
library(phylodyn)
selected_seasons <- c("2014-2015", "2015-2016", "2016-2017", "2017-2018", "2018-2019")

NY_ILI <- ilinet("state") %>%
  filter(region == "New York") %>%
  select("year", "week", "ilitotal", "week_start") %>%
  rename(ILI_total = ilitotal)

p_ILI <- visualize_flu_season(NY_ILI, time_series_names = "ILI_total", 
                              subset_seasons = selected_seasons, ncol = 5, season_label = F)
p_ILI
```

We can also visualize the inferred trajectory. 

```{r H3N2_tree}
H3N2_tree <- ape::read.nexus(system.file("extdata/H3N2/MCC.tree",
                                         package = "PILAF"))
H3N2_last_time <- 2019.0739726027398
ar2crw1_formula <- y ~ -1 + f(time, model = "ar", order = 2) +
      f(week, model = "rw1", cyclic = T, constr = F)
H3N2_Ne_forecast <- BNPR_forecast(H3N2_tree, last_time = H3N2_last_time, formula = ar2crw1_formula, pred = 0)

H3N2_Ne_data <- with(H3N2_Ne_forecast, 
                     data.frame(time = H3N2_last_time - result$summary.random$time$ID,
                                week = weeks,
                                H3N2_Ne = effpopmean,
                                lwr = effpop025,
                                upr = effpop975))
H3N2_Ne_data$year <- floor(H3N2_Ne_data$time)
p_H3N2_Ne <- visualize_flu_season(H3N2_Ne_data, time_series_names = "H3N2_Ne", 
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H3N2_Ne

H3N2_BNPR <- BNPR(H3N2_tree, lengthout = 300)
H3N2_BNPR_data <- with(H3N2_BNPR, 
                     data.frame(time = H3N2_last_time - result$summary.random$time$ID,
                                H3N2_Ne = effpopmean,
                                lwr = effpop025,
                                upr = effpop975))
H3N2_BNPR_data$week <- lubridate::week(lubridate::date_decimal(H3N2_BNPR_data$time))
H3N2_BNPR_data$year <- floor(H3N2_BNPR_data$time)
p_H3N2_BNPR <- visualize_flu_season(H3N2_BNPR_data, time_series_names = "H3N2_Ne", 
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H3N2_BNPR
```

```{r H1N1_tree}
H1N1_tree <- ape::read.nexus(system.file("extdata/H1N1/MCC.tree",
                                         package = "PILAF"))
H1N1_last_time <- 2019.0931506849315
ar2crw1_formula <- y ~ -1 + f(time, model = "ar", order = 2) +
      f(week, model = "rw1", cyclic = T, constr = F)
H1N1_Ne_forecast <- BNPR_forecast(H1N1_tree, last_time = H1N1_last_time, formula = ar2crw1_formula, pred = 0)
H1N1_Ne_data <- with(H1N1_Ne_forecast, data.frame(time = H1N1_last_time - result$summary.random$time$ID,
                                             week = weeks,
                                             H1N1_Ne = effpopmean,
                                             lwr = effpop025,
                                             upr = effpop975
                                             ))
H1N1_Ne_data$year <- floor(H1N1_Ne_data$time)
p_H1N1_Ne <- visualize_flu_season(H1N1_Ne_data, time_series_names = "H1N1_Ne",
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H1N1_Ne


H1N1_BNPR <- BNPR(H1N1_tree, lengthout = 300)
H1N1_BNPR_data <- with(H1N1_BNPR, 
                     data.frame(time = H1N1_last_time - result$summary.random$time$ID,
                                H1N1_Ne = effpopmean,
                                lwr = effpop025,
                                upr = effpop975))
H1N1_BNPR_data$week <- lubridate::week(lubridate::date_decimal(H1N1_BNPR_data$time))
H1N1_BNPR_data$year <- floor(H1N1_BNPR_data$time)
p_H1N1_BNPR <- visualize_flu_season(H1N1_BNPR_data, time_series_names = "H1N1_Ne", 
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_H1N1_BNPR
```

```{r}
Yamagata_tree <- ape::read.nexus(system.file("extdata/Yamagata/MCC.tree",
                                         package = "PILAF"))
Yamagata_last_time <- 2019.0739726027398
ar2crw1_formula <- y ~ -1 + f(time, model = "ar", order = 2) +
      f(week, model = "rw1", cyclic = T, constr = F)
Yamagata_Ne_forecast <- BNPR_forecast(Yamagata_tree, last_time = Yamagata_last_time, formula = ar2crw1_formula, pred = 0)
Yamagata_Ne_data <- with(Yamagata_Ne_forecast, data.frame(time = Yamagata_last_time - result$summary.random$time$ID,
                                             week = weeks,
                                             Yamagata_Ne = effpopmean,
                                             lwr = effpop025,
                                             upr = effpop975))

Yamagata_Ne_data$year <- floor(Yamagata_Ne_data$time)
p_Yamagata_Ne <- visualize_flu_season(Yamagata_Ne_data, time_series_names = "Yamagata_Ne",
                                  subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                  error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_Yamagata_Ne
```

```{r}
Victoria_tree <- ape::read.nexus(system.file("extdata/Victoria/MCC.tree",
                                         package = "PILAF"))
Victoria_last_time <- 2018.317808219178
ar2crw1_formula <- y ~ -1 + f(time, model = "ar", order = 2) +
      f(week, model = "rw1", cyclic = T, constr = F)
Victoria_Ne_forecast <- BNPR_forecast(Victoria_tree, last_time = Victoria_last_time, formula = ar2crw1_formula, pred = 0)
Victoria_Ne_data <- with(Victoria_Ne_forecast, data.frame(time = Victoria_last_time - result$summary.random$time$ID,
                                             week = weeks,
                                             Victoria_Ne = effpopmean,
                                             lwr = effpop025,
                                             upr = effpop975))
Victoria_Ne_data$year <- floor(Victoria_Ne_data$time)
p_Victoria_Ne <- visualize_flu_season(Victoria_Ne_data, time_series_names = "Victoria_Ne",
                                      subset_seasons = selected_seasons, ncol = 5, season_label = F,
                                      error_band_names = data.frame(lwr = "lwr", upr = "upr", stringsAsFactors = F))
p_Victoria_Ne
```

We can alsoview each season separately. 

```{r include = F}
theme_no_x <-  theme(
  axis.title.x = element_blank(), 
  axis.line.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.text.x = element_blank())
p_ILI <- p_ILI + ylab("ILI total") + theme_no_x
p_H3N2_Ne <- p_H3N2_Ne + ylab("H3N2 Ne") + theme_no_x + theme(strip.text = element_blank())
p_H1N1_Ne <- p_H1N1_Ne + ylab("H1N1 Ne") + theme(strip.text = element_blank())
p2 <- cowplot::plot_grid(p_ILI, p_H3N2_Ne, p_H1N1_Ne, ncol = 1, align = "v")
ggsave(p2, file = "../../projectingvirus/manuscript/Figures/data/Ne_ILI.pdf", width = 9, height = 5)

p_H3N2_BNPR <- p_H3N2_BNPR + ylab("H3N2 Ne") + theme_no_x + theme(strip.text = element_blank())
p_H1N1_BNPR <- p_H1N1_BNPR + ylab("H1N1 Ne") + theme(strip.text = element_blank())
p3 <- cowplot::plot_grid(p_ILI, p_H3N2_BNPR, p_H1N1_BNPR, ncol = 1, align = "v")
ggsave(p3, file = "../../projectingvirus/manuscript/Figures/data/Ne_ILI_BNPR.pdf", width = 9, height = 5)
```

cRW(1) + AR(2): Quantify the similarity between time series using Spearman correlation.

```{r}
joined_ts1 <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H3N2_Ne_data, year, week, H3N2_Ne), by = c("year", "week"))  %>%
  select(ILI_total, H3N2_Ne) 
cor.test(joined_ts1$ILI_total, joined_ts1$H3N2_Ne)
joined_ts2 <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H1N1_Ne_data, year, week, H1N1_Ne), by = c("year", "week")) %>%
  select(ILI_total, H1N1_Ne) 
cor.test(joined_ts2$ILI_total, joined_ts2$H1N1_Ne)
joined_all <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H3N2_Ne_data, year, week, H3N2_Ne), by = c("year", "week")) %>%
  dplyr::inner_join(select(H1N1_Ne_data, year, week, H1N1_Ne), by = c("year", "week")) %>%
  select(ILI_total, H3N2_Ne, H1N1_Ne)
lm_res <- lm(ILI_total ~ H3N2_Ne + H1N1_Ne, data = joined_all)
summary(lm_res)
```

BNPR: Regression analysis between ILI and the two virus time series.

```{r}
joined_ts1 <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H3N2_BNPR_data, year, week, H3N2_Ne), by = c("year", "week"))  %>%
  select(ILI_total, H3N2_Ne) 
cor.test(joined_ts1$ILI_total, joined_ts1$H3N2_Ne)
joined_ts2 <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H1N1_BNPR_data, year, week, H1N1_Ne), by = c("year", "week")) %>%
  select(ILI_total, H1N1_Ne) 
cor.test(joined_ts2$ILI_total, joined_ts2$H1N1_Ne)
joined_all <- select(NY_ILI, year, week, "ILI_total") %>%
  dplyr::inner_join(select(H3N2_BNPR_data, year, week, H3N2_Ne), by = c("year", "week")) %>%
  dplyr::inner_join(select(H1N1_BNPR_data, year, week, H1N1_Ne), by = c("year", "week")) %>%
  select(ILI_total, H3N2_Ne, H1N1_Ne)
lm_res <- lm(ILI_total ~ H3N2_Ne + H1N1_Ne, data = joined_all)
summary(lm_res)
```
