---
title: "Plot MCC tree"
author: "Meng Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot MCC tree}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
library(dplyr)
```

The document reproduced Figure in the manuscript. 

# H3N2

First, we load The maximum clade credibility tree (MCC) for H3N2 virus sampled from 2014 - 2019. A MCC tree is the tree in the posterior samples with the maximum sum of posterior clade probabilities. We 

```{r load_MCC_tree}
H3N2_MCC <- system.file("extdata/H3N2/MCC.tree", package = "PILAF")
H3N2_MCC_tree <- ape::read.nexus(H3N2_MCC)
H3N2_MCC_phylo <- phylodyn::summarize_phylo(H3N2_MCC_tree)
H3N2_last_time <- 2019.0739726027398 # last sampling time
H3N2_root_time = H3N2_last_time - max(H3N2_MCC_phylo$coal_times)
```

SkyGrid reconstruction was performed in BEAST and we can load the effective population size summary statistics as follows. 

```{r load_skygrid}
H3N2_skygrid <- system.file("extdata/H3N2/skygrid_reconstruction.tsv", package = "PILAF")
H3N2_skygrid <- readr::read_tsv(H3N2_skygrid, skip = 1) %>%
  mutate(Time = H3N2_last_time - Time) %>% 
  mutate(Label = "SkyGrid")
```

We can plot the MCC tree and inferred effective size trajectory as follows. 

```{r}
PILAF::plot_MCC_Ne(H3N2_MCC_tree, H3N2_root_time, H3N2_last_time, H3N2_skygrid, breaks = seq(2013, 2019))
```

We can also compare the results of SkyGrid with `BNPR` and `BNPR-PS` from `phylodyn`. 

```{r}
BNPR_traj <- phylodyn::BNPR(H3N2_MCC_phylo, lengthout = 300)
BNPR_PS_traj <- phylodyn::BNPR_PS(H3N2_MCC_phylo, lengthout = 300)
H3N2_all <- dplyr::bind_rows(H3N2_skygrid, 
                        BNPR_to_df(BNPR_traj, "BNPR", H3N2_last_time),
                        BNPR_to_df(BNPR_PS_traj, "BNPR PS", H3N2_last_time)) %>%
  mutate(Label = factor(Label, levels = c("SkyGrid", "BNPR", "BNPR PS")))
PILAF::plot_MCC_Ne(H3N2_MCC_tree, H3N2_root_time, H3N2_last_time, H3N2_all, main = "A/H3N2", breaks = seq(2013, 2019), c(1.3, 3))
```

# H1N1

The same process was repeated for H1N1. 

```{r H1N1}
H1N1_MCC <- system.file("extdata/H1N1/MCC.tree", package = "PILAF")
H1N1_MCC_tree <- ape::read.nexus(H1N1_MCC)
H1N1_MCC_phylo <- phylodyn::summarize_phylo(H1N1_MCC_tree)
H1N1_last_time <- 2019.0931506849315 # last sampling time
H1N1_root_time = H1N1_last_time - max(H1N1_MCC_phylo$coal_times)

H1N1_skygrid <- system.file("extdata/H1N1/skygrid_reconstruction.tsv", package = "PILAF")
H1N1_skygrid <- readr::read_tsv(H1N1_skygrid, skip = 1) %>%
  mutate(Time = H1N1_last_time - Time) %>% 
  mutate(Label = "SkyGrid")

BNPR_traj <- phylodyn::BNPR(H1N1_MCC_phylo, lengthout = 300)
BNPR_PS_traj <- phylodyn::BNPR_PS(H1N1_MCC_phylo, lengthout = 300)
H1N1_all <- dplyr::bind_rows(H1N1_skygrid, 
                        BNPR_to_df(BNPR_traj, "BNPR", H1N1_last_time),
                        BNPR_to_df(BNPR_PS_traj, "BNPR PS", H1N1_last_time)) %>%
  mutate(Label = factor(Label, levels = c("SkyGrid", "BNPR", "BNPR PS")))

PILAF::plot_MCC_Ne(H1N1_MCC_tree, H1N1_root_time, H1N1_last_time, H1N1_all, main = "A/H1N1", breaks = seq(2013, 2019), c(1, 3))
```

# Victoria

```{r Victoria}
Victoria_MCC <- system.file("extdata/Victoria/MCC.tree", package = "PILAF")
Victoria_MCC_tree <- ape::read.nexus(Victoria_MCC)
Victoria_MCC_phylo <- phylodyn::summarize_phylo(Victoria_MCC_tree)
Victoria_last_time <- 2018.317808219178 
Victoria_root_time = Victoria_last_time - max(Victoria_MCC_phylo$coal_times)

Victoria_skygrid <- system.file("extdata/Victoria/skygrid_reconstruction.tsv", package = "PILAF")
Victoria_skygrid <- readr::read_tsv(Victoria_skygrid, skip = 1) %>%
  mutate(Time = Victoria_last_time - Time) %>% 
  mutate(Label = "SkyGrid")

BNPR_traj <- phylodyn::BNPR(Victoria_MCC_phylo, lengthout = 300)
BNPR_PS_traj <- phylodyn::BNPR_PS(Victoria_MCC_phylo, lengthout = 300)
Victoria_all <- dplyr::bind_rows(Victoria_skygrid, 
                        BNPR_to_df(BNPR_traj, "BNPR", Victoria_last_time),
                        BNPR_to_df(BNPR_PS_traj, "BNPR PS", Victoria_last_time)) %>%
  mutate(Label = factor(Label, levels = c("SkyGrid", "BNPR", "BNPR PS")))
PILAF::plot_MCC_Ne(Victoria_MCC_tree, Victoria_root_time, Victoria_last_time, Victoria_all, main = "B/Victoria", breaks = seq(2013, 2019), c(1, 3))
```

# Yamagata

```{r Yamagata}
Yamagata_MCC <- system.file("extdata/Yamagata/MCC.tree", package = "PILAF")
Yamagata_MCC_tree <- ape::read.nexus(Yamagata_MCC)
Yamagata_MCC_phylo <- phylodyn::summarize_phylo(Yamagata_MCC_tree)
Yamagata_last_time <- 2019.0739726027398 
Yamagata_root_time = Yamagata_last_time - max(Yamagata_MCC_phylo$coal_times)

Yamagata_skygrid <- system.file("extdata/Yamagata/skygrid_reconstruction.tsv", package = "PILAF")
Yamagata_skygrid <- readr::read_tsv(Yamagata_skygrid, skip = 1) %>%
  mutate(Time = Yamagata_last_time - Time) %>% 
  mutate(Label = "SkyGrid")

BNPR_traj <- phylodyn::BNPR(Yamagata_MCC_phylo, lengthout = 300)
BNPR_PS_traj <- phylodyn::BNPR_PS(Yamagata_MCC_phylo, lengthout = 300)
Yamagata_all <- dplyr::bind_rows(Yamagata_skygrid, 
                        BNPR_to_df(BNPR_traj, "BNPR", Yamagata_last_time),
                        BNPR_to_df(BNPR_PS_traj, "BNPR PS", Yamagata_last_time)) %>%
  mutate(Label = factor(Label, levels = c("SkyGrid", "BNPR", "BNPR PS")))
PILAF::plot_MCC_Ne(Yamagata_MCC_tree, Yamagata_root_time, Yamagata_last_time, Yamagata_all, main = "B/Yamagata", breaks = seq(2013, 2019), c(1, 3))
```

```{r}
ylimits <- c(0.5e-2, 3e2)
pdf("~/Desktop/H3N2.pdf", width = 4, height = 8)
PILAF::plot_MCC_Ne(H3N2_MCC_tree, H3N2_root_time, H3N2_last_time, H3N2_all, main = "A/H3N2", breaks = seq(2013, 2019, by = 2), c(1.3, 3), ylimits = ylimits)
dev.off()
pdf("~/Desktop/H1N1.pdf", width = 4, height = 8)
PILAF::plot_MCC_Ne(H1N1_MCC_tree, H1N1_root_time, H1N1_last_time, H1N1_all, main = "A/H1N1", breaks = seq(2013, 2019, by = 2), c(1.3, 3), ylimits = ylimits)
dev.off()
pdf("~/Desktop/Victoria.pdf", width = 4, height = 8)
PILAF::plot_MCC_Ne(Victoria_MCC_tree, Victoria_root_time, Victoria_last_time, Victoria_all, main = "B/Victoria", breaks = seq(2014, 2018), c(1.3, 3), ylimits = ylimits)
dev.off()
pdf("~/Desktop/Yamagata.pdf", width = 4, height = 8)
PILAF::plot_MCC_Ne(Yamagata_MCC_tree, Yamagata_root_time, Yamagata_last_time, Yamagata_all, main = "B/Yamagata", breaks = seq(2014, 2019), c(1.3, 3), mask_range = c(2018.3, 2020), ylimits = ylimits)
dev.off()
```
